Concentration of Maxima and Fundamental Limits in High-Dimensional Testing and Inference

### 2020-12-22 1.intro/introduction.tex
Addressed the minor typos of R2
I defined the command "\stilian" in the preamble and used it to make changes in response to the minor typos.

Please see reviews/GS_Annotated.pdf.

Model \eqref{eq:model-additive} is simple yet ubiquitous.
Consider the linear model
\begin{equation*}
 Y = X\mu + \xi,
\end{equation*}
where $\mu$ is a $p$-dimensional vector of regression coefficients of interest to be inferred from observations of $X$ and $Y$.
If the design matrix $X$ is of full column rank, then the ordinary least squares (OLS) estimator of $\mu$ can be formed 
\begin{equation} \label{eq:OLS}
    \widehat{\mu} = \left(X'X\right)^{-1}X'Y = \mu + \epsilon,
\end{equation}
where $\epsilon := (X'X)^{-1}X'\xi$.
Hence we recover the generic problem \eqref{eq:model-additive}. 
Signal detection is therefore equivalent to the problem of testing the global null model, and support recovery problem corresponds to the fundamental problem of variable selection.

>>>>

Model \eqref{eq:model-additive} is simple yet ubiquitous.
Consider the linear model
\begin{equation*}
 Y = X\mu + \xi,
\end{equation*}
where $\mu$ is a $p$-dimensional vector of regression coefficients of interest to be inferred from observations of $X$ and $Y$.
If the design matrix $X$ is of full column rank, then the ordinary least squares (OLS) estimator of $\mu$ can be formed 
\begin{equation} \label{eq:OLS}
    \widehat{\mu} = \left(X'X\right)^{-1}X'Y = \mu + \epsilon,
\end{equation}
where $\epsilon := (X'X)^{-1}X'\xi$. \stilian{R2, Typo 2}{I see no problems here except that we are not in the high-dimensional regime $p\gg n$. If we were to address that, we need to expand the discussion. 1) Talk about sparsity of $\mu$ 2) Introduce the assumption that if ${\rm supp}(\mu) = r <n$, then every $r$ of the $p$ columns of 
$X$ must be linearly independent -- is there a name for such a condition in the literature? We know it happens with probability one if $X$ has independent, say Gaussian, entries.  3) Then, it seems clear that the support of $\mu$ will be identifiable if we replace the inverse with a Moore-Penrose inverse. BTW, an intriguing question: Given a design matrix $X$, consider the quantity:
$$
c_r(X) :=\min_{w\subset \{1,\cdots,p\},\ |w| = r} {\| (X_w' X_w)^{-1}\| \over \| X_w'X_w\| },
$$
where the minimum is taken over all subsets of indices $w$ with cardinality $r<n$ and $X_w:= (x_{i,j})_{j\in w}$ is the sub-matrix of $X$ obtained
by selecting the columns indexed by the elements in $w\subset\{1,\cdots,p\}$.  Intuitively, $c_r(X)$ gives us the overall difficulty of identifying a 
sparse signal $\mu$ with support of size $r$. \fbox{Going off on a tangent here...}}
Hence we recover the generic problem \eqref{eq:model-additive}. 
Signal detection is therefore equivalent to the problem of testing the global null model, and support recovery problem corresponds to the fundamental problem of variable selection.

### 2020-12-22 2.background/2.Exercises.tex
Update 2.Exercises.tex
Addressed the minor typos of R2
I defined the command "\stilian" in the preamble and used it to make changes in response to the minor typos.

Please see reviews/GS_Annotated.pdf.

\fbox{ TO DO: Verify and add}

@@

 \item (Inspired by \cite{resnick1973almost}) Let $\epsilon(i)\sim F,\ i=1,2,\cdots$ have arbitrary dependence 
 and $u_p = F^{\leftarrow}(1-1/p)$.

@@

{\em Hint:} \fbox{ Give a good hint. }

\item  
\end{enumerate}

>>>>

\stilian{R2, Typo 4}{I suggest we move all exercises to the online Supplement.  The supplement can have several chapters, the first of which could be 
"A crash course on extremes" or ``Selected background from EVT".  It could address some comments on the reviewers that the practitioners could need 
a lot of ``hand holding".  Similarly, in the Supplement we can put other ``basic" things we would like to have like ``Standard" derivations on the Sidak, BH procedure
the definition of Tukey's Higher Criticism statistics and the results from Empirical processes that are used to study it...} \fbox{ TO DO: Verify and add}

@@

 \item (Inspired by \cite{resnick1973almost}) Let $\epsilon(i)\sim F,\ i=1,2,\cdots$ have arbitrary dependence 
 and $u_p = F^{\leftarrow}(1-1/p) \to \infty$.

@@

\stilian{R2, Typo 5}{ \fbox{TO DO} {\em Hint:} \fbox{ Give a good hint. } }

\end{enumerate}

### 2020-12-22 2.background/2.statistical_procedures.tex
Addressed the minor typos of R2
I defined the command "\stilian" in the preamble and used it to make changes in response to the minor typos.

Please see reviews/GS_Annotated.pdf.

Examples of thresholding procedures include ones that aim to control FWER \eqref{eq:FWER-FWNR} --- Bonferroni's \cite{dunn1961multiple}, Sid\'ak's \citep{vsidak1967rectangular}, Holm's \citep{holm1979simple}, and Hochberg's procedure \citep{hochberg1988sharper} --- as well as procedures that target FDR \eqref{eq:FDR-FNR}, such as the Benjamini-Hochberg \cite{benjamini1995controlling} and the Cand\'es-Barber procedure \cite{barber2015controlling}.
Indeed, the class of thresholding procedures \eqref{eq:thresholding-procedure} is so general that it contains most (but not all) of the statistical procedures in the multiple testing literature.
% \cite{roquain2011type}.

@@

Recall the order statistics of our observations $x_{[1]} \ge x_{[2]}  \ge \ldots \ge x_{[p]}$.
\begin{definition}[Benjamini-Hochberg's procedure]
Let $i^*$ be the largest index such that
$$
\overline{F}(x_{[i]}) \le \alpha i/p.
$$
The Benjamini-Hochberg procedure with level $\alpha$ is the thresholding procedure with threshold
\begin{equation} \label{eq:BH-procedure}
    t_p(x) = x_{[i^*]},
\end{equation}
\end{definition}
The \ac{BH} procedure is shown to control the FDR at level $\alpha$ when the $x(i)$'s are independent \citep{benjamini1995controlling}.
Variations of the \ac{BH} procedure have been proposed to control the \ac{FDR} under dependent observations \citep{benjamini2001control}.

>>>>

Examples of thresholding procedures include ones that aim to control FWER \eqref{eq:FWER-FWNR} --- Bonferroni's \cite{dunn1961multiple}, Sid\'ak's \citep{vsidak1967rectangular}, Holm's \citep{holm1979simple}, and Hochberg's procedure \citep{hochberg1988sharper} --- as well as procedures that target FDR \eqref{eq:FDR-FNR}, such as the Benjamini-Hochberg \cite{benjamini1995controlling} and the \stilian{R2,Typo 3}{fixed} Cand\`es-Barber procedure \cite{barber2015controlling}.
Indeed, the class of thresholding procedures \eqref{eq:thresholding-procedure} is so general that it contains most (but not all) of the statistical procedures in the multiple testing literature.
% \cite{roquain2011type}.

@@

Recall the order statistics of our observations $x_{[1]} \ge x_{[2]}  \ge \ldots \ge x_{[p]}$.
\begin{definition}[Benjamini-Hochberg's procedure] \label{def:BH}
Let $i^*$ be the largest index such that
$$
\overline{F}(x_{[i]}) \le \alpha i/p.
$$
The Benjamini-Hochberg procedure with level $\alpha$ is the thresholding procedure with threshold
\begin{equation} \label{eq:BH-procedure}
    t_p(x) = x_{[i^*]},
\end{equation}
\end{definition}
The \ac{BH} procedure is shown to control the FDR at level $\alpha$ when the $x(i)$'s are independent \citep{benjamini1995controlling}.
Variations of the \ac{BH} procedure have been proposed to control the \ac{FDR} under dependent observations \citep{benjamini2001control}.

### 2020-12-22 3.phase_transitions/3.global-tests.tex
Addressed the minor typos of R2
I defined the command "\stilian" in the preamble and used it to make changes in response to the minor typos.

Please see reviews/GS_Annotated.pdf.

To the best of our knowledge, performance of simple statistics such as $L_1$, $L_2$ norms, and $S$ in this weak signal setting have not been reported in the literature perhaps due to a perceived lack of novelty.
Our first Theorem investigates the performance of these routine statistics for detecting sparse signals in high-dimensions, and summarizes the known results.

>>>>

To the best of our knowledge, performance of simple statistics such as $L_1$, $L_2$ norms, and \stilian{R2, Typo 7.  I suggest writing the following in red, but note that
``S" can be confused with the support.  Can we use $\Sigma$ or ${\rm Sum}$? }{the sum statistic $S$ in \eqref{eq:sum-statistic}}
in this weak signal setting have not been reported in the literature perhaps due to a perceived lack of novelty.
Our first \stilian{Suggest replacing ``Theorem" with ``theorem" and using a capital letter only when we quote a specific Theorem X.Y. -- same for Lemma, Proposition, etc.}{Theorem} investigates the performance of these routine statistics for detecting sparse signals in high-dimensions, and summarizes the known results.

### 2020-12-22 4.exact_support_recovery/exact_support_recovery.tex
minor edits
added

1) ?.Example.tex files for each chapter conforming to the current style.

2) Began the "index".  If you want to include a term do \index{term}{term}

3) Fixed the level of the table-of-contents.
Addressed the minor typos of R2
I defined the command "\stilian" in the preamble and used it to make changes in response to the minor typos.

Please see reviews/GS_Annotated.pdf.

Unfortunately, as pointed out in Section \ref{sec:risks-relations}, vanishing Hamming loss is only sufficient, not necessary for support recovery \eqref{eq:exact-recovery}, and the sharp result does not carry over directly to the study of the probability of support recovery or exact support recovery risk.
More importantly, despite their elegance, these Hamming loss-minimax studies naturally amounts to the analysis of the elementary case of iid data, and is by design blind to non-trivial error-dependence structures.
This prevents us from fully exploring of the phase transition phenomena under other dependence conditions.
% This is arguably a consequence of the inherent limitation of the Hamming-loss approach. 
% For practitioners, a minimax statement in the Gaussian case seems to offer little guidance.
% Indeed, in applications, independence is an exception rather than the rule; Gaussianity of the errors may also be unnecessarily restrictive.

@@

>>>>

Unfortunately, as pointed out in Section \ref{sec:risks-relations}, vanishing Hamming loss is only sufficient, not necessary for support recovery \eqref{eq:exact-recovery}, and the sharp result does not carry over directly to the study of the probability of support recovery or exact support recovery risk.
More importantly, despite their elegance, these Hamming loss-minimax studies naturally \stilian{replace ``amounts" by ``reduce"}{amounts} to the analysis of the elementary case of iid data, and is by design blind to non-trivial error-dependence structures.
This prevents us from fully exploring of the phase transition phenomena under other dependence conditions.
% This is arguably a consequence of the inherent limitation of the Hamming-loss approach. 
% For practitioners, a minimax statement in the Gaussian case seems to offer little guidance.
% Indeed, in applications, independence is an exception rather than the rule; Gaussianity of the errors may also be unnecessarily restrictive.

@@

\input{4.exact_support_recovery/4.Exercises.tex}

### 2020-12-22 5.URS/5.proofs-UDD-URS.tex
Addressed the minor typos of R2
I defined the command "\stilian" in the preamble and used it to make changes in response to the minor typos.

Please see reviews/GS_Annotated.pdf.

\begin{proof}[Proof of Proposition \ref{prop:Gaussian-maxima-expectation-lower-bound}]
% Assume without loss of generality that the $\epsilon(j)$'s have unit variances.
Define the canonical (pseudo) metric on $S_p$,
$$
d(i,j) = \sqrt{\E\left[\left(\epsilon(i)-\epsilon(j)\right)^2\right]}.
$$
It can be easily checked that the canonical metric takes values between 0 and 2.
For arbitrary $\delta\in(0,1)$, take $\gamma = \sqrt{2(1-\delta)}$, and
let $\mathcal{N}$ be a $\gamma$-packing of $S_p$. That is, let $\mathcal{N}$ be a subset of $S_p$, such that for any $i,j\in\mathcal{N}$, $i\neq j$, we have $d(i,j)\ge\gamma$, i.e.,
\begin{equation}\label{e:gamma-packing-def}
d(i,j) = \sqrt{2\left(1-\Sigma_p(i,j)\right)} \ge \gamma = \sqrt{2(1-\delta)},
\end{equation}
or equivalently, $\Sigma_p(i,j) \le \delta$.
We claim that we can find a $\gamma$-packing $\mathcal{N}$ whose number of elements is at least 
\begin{equation} \label{eq:packing-number-lower-bound}
    |\mathcal{N}| \ge q/N(\delta).
\end{equation}
Indeed, $\mathcal{N}$ can be constructed iteratively as follows:
\begin{itemize}[
    align=left,
    leftmargin=4em,
    itemindent=0pt,
    labelsep=0pt,
    labelwidth=4em
    ]
    \raggedright
    \item[{\bf Step 1:}] Set $S_p^{(1)}:=S_p$ and $\mathcal{N}:=\{j_1\}$, where $j_1\in S_p^{(1)}$ is an arbitrary element. Set $k:=1$.\\
    \item[{\bf Step 2:}] Set $S_p^{(k+1)}:=S_p^{(k)}\setminus B_\gamma(j_k)$, where
    $$
    B_\gamma(j_k) := \{i\in S_p: \;d(i,j_k) < \gamma \equiv \sqrt{2(1-\delta)}\}.
    $$
    \item[{\bf Step 3:}] If $S_p^{(k)} \neq \emptyset$, pick an arbitrary $j_{k+1}\in S_p^{(k)}$, set $\mathcal{N}:=\mathcal{N}\cup\{j_{k+1}\}$, and $k:=k+1$, go to step 2; otherwise, stop.
\end{itemize}
By the definition of UDD (see Definition \ref{def:UDD}), there are at most $N(\delta)$ coordinates whose covariance with $\epsilon(j)$ exceed $\delta$. 
Therefore at each iteration, $\left|B_\gamma(j_k)\right|\le N(\delta)$, and hence
$$
\left|S_p^{(k+1)}\right| \ge \left|S_p^{(k)}\right| - \left| B_\gamma(j_k)\right| \ge q - kN(\delta).
$$
The construction can continue for at least $q/N(\delta)$ iterations, and we have $|\mathcal{N}| \ge \lfloor q/N(\delta) \rfloor$ as desired.

>>>>

\begin{proof}[Proof of Proposition \ref{prop:Gaussian-maxima-expectation-lower-bound}]
% Assume without loss of generality that the $\epsilon(j)$'s have unit variances.
Define the canonical (pseudo) metric on $S_p$,
$$
d(i,j) = \sqrt{\E\left[\left(\epsilon(i)-\epsilon(j)\right)^2\right]}.
$$
It can be easily checked that the canonical metric takes values between 0 and 2.
For arbitrary $\delta\in(0,1)$, take $\gamma = \sqrt{2(1-\delta)}$, and
let $\mathcal{N}$ be a $\gamma$-packing of $S_p$. That is, let $\mathcal{N}$ be a subset of $S_p$, such that for any $i,j\in\mathcal{N}$, $i\neq j$, we have $d(i,j)\ge\gamma$, i.e.,
\begin{equation}\label{e:gamma-packing-def}
d(i,j) = \sqrt{2\left(1-\Sigma_p(i,j)\right)} \ge \gamma = \sqrt{2(1-\delta)},
\end{equation}
or equivalently, $\Sigma_p(i,j) \le \delta$.
We claim that we can find a $\gamma$-packing $\mathcal{N}$ whose number of elements is at least 
\begin{equation} \label{eq:packing-number-lower-bound}
    |\mathcal{N}| \ge q/N(\delta).
\end{equation}
Indeed, $\mathcal{N}$ can be constructed iteratively as follows:
\begin{itemize}    \raggedright
    \item[{\bf 1:}] Set $S_p^{(1)}:=S_p$ and $\mathcal{N}:=\{j_1\}$, where $j_1\in S_p^{(1)}$ is an arbitrary element. Set $k:=1$.\\
    \item[{\bf 2:}] Set $S_p^{(k+1)}:=S_p^{(k)}\setminus B_\gamma(j_k)$, where
    $$
    B_\gamma(j_k) := \{i\in S_p: \;d(i,j_k) < \gamma \equiv \sqrt{2(1-\delta)}\}.
    $$
    \item[{\bf 3:}] If $S_p^{(k)} \neq \emptyset$, pick an arbitrary $j_{k+1}\in S_p^{(k)}$, set $\mathcal{N}:=\mathcal{N}\cup\{j_{k+1}\}$, and $k:=k+1$, go to step 2; otherwise, stop.
\end{itemize}
By the definition of UDD (see Definition \ref{def:UDD}), there are at most $N(\delta)$ coordinates whose covariance with $\epsilon(j)$ exceed $\delta$. 
Therefore at each iteration, $\left|B_\gamma(j_k)\right|\le N(\delta)$, and hence
$$
\left|S_p^{(k+1)}\right| \ge \left|S_p^{(k)}\right| - \left| B_\gamma(j_k)\right| \ge q - kN(\delta).
$$
The construction can continue for at least $q/N(\delta)$ iterations, and we have $|\mathcal{N}| \ge \lfloor q/N(\delta) \rfloor$ as desired.

### 2020-12-22 6.GWAS/6.chi-square.tex
Addressed the minor typos of R2
I defined the command "\stilian" in the preamble and used it to make changes in response to the minor typos.

Please see reviews/GS_Annotated.pdf.

\begin{theorem} \label{thm:chi-squared-approx-boundary}
Consider the high-dimensional chi-squared model \eqref{eq:model-chisq} with signal sparsity and size as described in \eqref{eq:signal-sparsity} and \eqref{eq:signal-size}.
The function 
\begin{equation} \label{eq:approx-boundary-chisquared}
    h(\beta) = \beta
\end{equation}
characterizes the phase transition of approximate support recovery problem.
Specifically, if $\underline{r} > {h}(\beta)$, then the \ac{BH}procedure $\widehat{S}_p$ (defined in Section \ref{sec:statistical-procedures}) with slowly vanishing (see Definition \ref{def:slowly-vanishing}) nominal FDR levels achieves asymptotically approximate support recovery in the sense of \eqref{eq:support-recovery-success}.

@@

As discussed in Section \ref{sec:motivation-chisq}, support recovery problems in the chi-square model with $\nu=1$, , corresponds to the support recovery problems in the additive model under two-sided alternatives.
This implies that the asymptotic signal size requirements are identical between the two-sided alternative and its one-sided counterpart, in order to achieve exact support recovery.
As we shall see in numerical experiments (in Section \ref{sec:numerical} below), the difference is not very pronounced even in moderate dimensions, and vanishes as $p\to\infty$, in accordance with Theorem \ref{thm:chi-squared-exact-boundary}.

>>>>

\begin{theorem} \label{thm:chi-squared-approx-boundary}
Consider the high-dimensional chi-squared model \eqref{eq:model-chisq} with signal sparsity and size as described in \eqref{eq:signal-sparsity} and \eqref{eq:signal-size}.
The function 
\begin{equation} \label{eq:approx-boundary-chisquared}
    h(\beta) = \beta
\end{equation}
characterizes the phase transition of approximate support recovery problem.
Specifically, if $\underline{r} > {h}(\beta)$, then the \ac{BH} procedure $\widehat{S}_p$ (defined in Section \ref{sec:statistical-procedures}) with slowly vanishing (see Definition \ref{def:slowly-vanishing}) nominal FDR levels achieves asymptotically approximate support recovery in the sense of \eqref{eq:support-recovery-success}.

@@

As discussed in Section \ref{sec:motivation-chisq}, support recovery problems in the chi-square model with $\nu=1$ correspond to the support recovery problems in 
the additive model under two-sided alternatives. This implies that the asymptotic signal size requirements are identical between the two-sided alternative and its 
one-sided counterpart, in order to achieve exact support recovery. As we shall see in numerical experiments (in Section \ref{sec:numerical} below), the difference 
is not very pronounced even in moderate dimensions, and vanishes as $p\to\infty$, in accordance with Theorem \ref{thm:chi-squared-exact-boundary}.

### 2020-12-22 6.GWAS/6.optimal-design.tex
Addressed the minor typos of R2
I defined the command "\stilian" in the preamble and used it to make changes in response to the minor typos.

Please see reviews/GS_Annotated.pdf.

Fortunately, the conditional distributions of genotypes in the healthy control groups are often estimated by existing studies, and are made available by consortia such as the NHGRI-EBI GWAS catalog \cite{macarthur2016new}.
% Assume (after appropriate relabelling, hence without loss of generality) that the first variant is associated with an increased risk of disease, and is henceforth referred to as the risk variant.
We denote the conditional frequency of the first genetic variant in the control group as $(f, 1-f)$, where
\begin{equation} \label{eq:RAF}
    f := \mu_{21} / \phi_2.
\end{equation}
The multinomial probability is fully parametrized by the new trio: $(f, \phi_1, R)$.
\begin{center}
    \begin{tabular}{cccc}
    \hline
    & \multicolumn{2}{c}{Genotype} \\
    \cline{2-3}
    Probabilities & Variant 1 & Variant 2 & Total by phenotype \\
    \hline
    Cases & $\frac{\phi_1fR}{fR+1-f}$ & $\frac{\phi_1(1-f)}{fR+1-f}$ & $\phi_1$ \\
    Controls & $f(1-\phi_1)$ & $(1-f)(1-\phi_1)$ & $1-\phi_1$ \\
    \hline
    \end{tabular}
\end{center}
Proposition \ref{prop:signal-size-odds-ratio} may also be re-stated in terms of the new parametrization.

>>>>

Fortunately, the conditional distributions of genotypes in the healthy control groups are often estimated by existing studies, and are made available by consortia such as the NHGRI-EBI GWAS catalog \citep{macarthur2016new}.
% Assume (after appropriate relabelling, hence without loss of generality) that the first variant is associated with an increased risk of disease, and is henceforth referred to as the risk variant.
We denote the conditional frequency of the first genetic variant in the control group as $(f, 1-f)$, where
\begin{equation} \label{eq:RAF}
    f := \mu_{21} / \phi_2.
\end{equation}
The multinomial probability is fully parametrized by the new trio: $(f, \phi_1, R)$.
\begin{center}
    \begin{tabular}{cccc}
    \hline
    & \multicolumn{2}{c}{Genotype} \\
    \cline{2-3}
    Probabilities & Variant 1 & Variant 2 & Total by phenotype \\
    \hline
    Cases & $\frac{\phi_1fR}{fR+1-f}$ & $\frac{\phi_1(1-f)}{fR+1-f}$ & $\phi_1$ \\
    Controls & $f(1-\phi_1)$ & $(1-f)(1-\phi_1)$ & $1-\phi_1$ \\
    \hline
    \end{tabular}
\end{center}
Proposition \ref{prop:signal-size-odds-ratio} may also be re-stated in terms of the new parametrization.

### 2020-12-22 preamble.tex
Addressed the minor typos of R2
I defined the command "\stilian" in the preamble and used it to make changes in response to the minor typos.

Please see reviews/GS_Annotated.pdf.

{\bf Acknowledgements.} The authors gratefully acknowledge the support of their families and all colleagues from the Statistics Department 
at the University of Michigan, Ann Arbor.   Special thanks (in alphabetical order) go to Xuming He, Tailen Hsing, Michalis Kallitsis, Liza Levina, 
Rodderick Little, Ya'acov Ritov, Kerby Shedden, Jinqi Shen, Jonathan Terhorst, Gongjun Xu, \fbox{add more}
The authors were partially supported by the NSF program {\em Algorithms for Threat Detection}.

>>>>

{\bf Acknowledgements.} The authors gratefully acknowledge the support of their families and all colleagues from the Statistics Department 
at the University of Michigan, Ann Arbor.   Special thanks (in alphabetical order) go to Xuming He, Tailen Hsing, Michalis Kallitsis, Liza Levina, 
Rodderick Little, Ya'acov Ritov, Kerby Shedden, Jinqi Shen, Jonathan Terhorst, Gongjun Xu, \stilian{R2, Typo 1}{\fbox{add more}}
The authors were partially supported by the NSF program {\em Algorithms for Threat Detection}.
